Trafficking synaptic cargo involves tradeoffs in speed and precision

### 2015-12-09 latex/paper-v2.tex

\begin{figure}[!tb]
\begin{center}
\includegraphics[width=0.9\columnwidth]{04_sushi_belt.png}
\caption{Multiple strategies for transport in a model including nonuniform microtubular detachment/activation.
\textbf{(A)} Schematic of microtubular transport model with irreversible detachment in a branched morphology. The localized degradation reactions ($u^{\star}_i \xrightarrow{d_i} \varnothing$) is omitted.
\textbf{(B)} Multiple strategies for producing a desired distribution of activated biomolecules ($\tilde{u}^{\star}$). When the timescale of detachment/delivery is sufficiently slow, the distribution of cargo on the microtubules approaches a quasi-steady-state (transport step). This known distribution can then be transformed into the desired distribution for $\tilde{u}$ (detachment step). As long as these two steps are appropriately matched (blends of blue), then the desired distribution will be achieved (CA1 cell, right).
\textbf{(C-E)} Normalized distributions of $u$ (red), $u^\star$ (blue), and $\tilde{u}^\star$ (black) as $t \rightarrow \infty$ for various strategies diagrammed in panel B (see circled red numbers). }
\end{center}
\end{figure}

@@

We examined this possibility in a simpler model --- an unbranched cable --- so that we could develop simple heuristics that precisely achieve a desired distribution of cargo while minimizing the convergence time.
In the reduced model, cargo begins on the left end of the cable and is transported to a number of synaptic sites, each of which is modeled as a double-exponential curve.

\begin{figure}[!tb]
\begin{center}
\includegraphics[width=0.9\columnwidth]{06_heterosynaptic_cable.png}
\caption{Producing speed and specificity for particular stimulation patterns does not extend to other patterns. Cargo begins on the left end and is transported to the right.
\textbf{(A-C)} Steady-state cargo profiles (red) for three stimulation patterns (black arrows) along an unbranched cable. The dotted black line corresponds to the `target' steady-state level at each synapse.
\textbf{(A)} When the timescale of detachment is sufficiently slow, cargo can be evenly distributed to the synapses regardless of their number and position. Transport parameters were set according to the procedure shown in figure 4D.
\textbf{(B)} When detachment is n\"iavely increased (all rates multiplicatively scaled) a proximal bias in the steady-state distribution of cargo across all stimulation patterns.
\textbf{(C)} Transport rate constants, $a_i$ and $b_i$, were tuned to optimize the distribution of cargo to six equally spaced synapses (top row); detachment rate constants were the same as in panel B. Changing the number of synapses (middle row) or the position of the synapses (bottom row) causes the unequal distribution of cargo to synapses.
\textbf{(D)} Tradeoff curves between non-specificity and convergence rate for six evenly spaced synapses (top row of A-C). The curve shifts to the left and becomes non-monotonic as the anterograde bias ($\beta$) increases.
\textbf{(E)} Tradeoff curves for two levels of anterograde bias on multiple simulations; for each simulation six synapse positions were drawn from random uniform distribution over the length of the cable.
\textbf{(F)} Same as panel E, except each the number of synapses was randomly chosen between 3 to 9 for each simulation.
}
\end{center}
\end{figure}

@@

We then hand-tuned the transport rate constants to deliver equal cargo to six evenly spaced synaptic sites (top row of Fig. 6C).
Specifically, we increased the anterograde rate constants ($a_i$) and decreased the retrograde rate constants ($b_i$) by a decreasing linear function of position so that $a_{N-1} = b_{N-1}$ at the right side of the cable.
On the left end of the cable, we set $a_1 = 0.5 + \beta$ and $b_1 = 0.5 - \beta$, where $\beta$ is the parameter controlling anterograde bias
Intuitively, this pushes cargo past the proximal synaptic sites more quickly, canceling out the exponential delivery bias induced by fast detachment.

@@

All simulation code is available on: https://github.com/ahwillia/Williams-etal-Synaptic-Transport

\subsection*{Model of single-particle transport}

Let $x_n$ denote the position of a particle along a 1-dimensional cable at timestep $n$. Let $v_n$ denote the velocity of the particle at timestep $n$; for simplicity, we assume the velocity can take on three discrete values, $v_n = \{-1,0,1\}$, corresponding to a discrete retrograde movement, pause, or anterograde movement. As a result, $x_n$ is constrained to take on integer values. In the memoryless transport model (top plots in Fig. 1B, 1D, and 1F), we assume that $v_n$ is drawn with fixed probabilities on each step. The update rule for position is:
$$x_{n+1} = x_n + v_n$$
$$v_{n+1} = \left\{\begin{matrix}
-1 & \text{with probability } p_-\\ 
0 & \text{with probability } p_0\\  
1 & \text{with probability } p_+
\end{matrix}\right.$$

We chose $p_- = 0.2$, $p_0 = 0.35$ and $p_+ = 0.45$ for the illustration shown in Figure 1. For the model with history-dependence (bottom plots in Fig. 1B, 1D, and 1F), the movement probabilites at each step depend on the previous movement. For example, if the motor was moving in an anterograde direction on the previous timestep, then it is more likely to continue to moving in that direction in the next time step. In this case the update rule utilizes \textit{conditional} probabilities:
$$v_{n+1} = \left\{\begin{matrix}
-1 & \text{with probability } p(-|v_n)\\ 
0 & \text{with probability } p(0|v_n)\\  
1 & \text{with probability } p(+|v_n)
\end{matrix}\right.$$

In the extreme case of history-dependence, the particle always steps in the same direction as the previous time step.
\begin{center}
  \begin{tabular}{ l | c  c  c }
    & $v_n=-1$ & $v_n=0$ & $v_n=1$ \\
    \hline
    $p(v_{n+1} = -1)$ & $1$ & $0$ & $0$ \\ 
    $p(v_{n+1} = 0)$  & $0$ & $1$ & $0$ \\ 
    $p(v_{n+1} = 1)$  & $0$ & $0$ & $1$ \\
    \hline
  \end{tabular}
\end{center}
We introduce a parameter $k \in [0,1]$ to linearly interpolate between this extreme case and the memoryless model. The rest of the panels in the figure show individual runs, 5000 molecule simulations, and mass action approximations for $k=0.5$.

\subsection*{Model of bulk/mass-action transport}
Transport in an unbranched cable (equation 1) leads to the following system of differential equations:
$$
\begin{array}{lcl}
\dot{u}_1 & = & -a_1 u_1 + b_1 u_2 \\
\dot{u}_2 & = & a_1 u_1 - (a_2+b_1) u_2 + b_2 u_3 \\
~ & \vdots & ~ \\
\dot{u}_i & = & a_{i-1} u_{i-1} - (a_i+b_{i-1}) u_i + b_i u_{i+1} \\
~ & \vdots & ~ \\
\dot{u}_N & = & a_{N-1} u_{N-1} - b_{N-1} u_N 
\end{array}
$$
Similar systems result from branched mass-action models. The steady-state ratio of cargo in two neighboring compartments equals the ratio of transport rate constants (equation 2). This can be immediately verified at either end of the cable:
$$
\dot{u}_1 = 0 ~\Rightarrow~ -a_{1} u_{1} + b_{1} u_2 = 0 ~\Rightarrow~ \frac{u_1}{u_2} \Bigg |_{ss} = \frac{b_1}{a_1}
$$
Successively applying the same logic down the cable confirms the condition in equation 2 holds globally. A similar procedure (starting at the tips and moving in) can be applied to branched morphologies to prove the more general condition (equation 4).

% The same result can be shown in a branched tree morphology. Here, each compartment has exactly one parent, but can have multiple children at branch points. Consider a parent-child pair of compartments, $u_p$ and $u_c$, with anterograde/retrograde transport rate constants $a$ and $b$. At steady-state, there is no net flux between $u_c$ and its children:
% $$
% \dot{u}_c = a u_p - b u_c + \underbrace{\sum_j (b_j u_j - a_j u_c)}_{= 0 \text{ , at ss}}
% $$
% Thus, we have:
% $$
% \dot{u}_c = a u_p - b u_c = 0 ~\Rightarrow~ \frac{u_p}{u_c} \Bigg |_{ss} = \frac{b}{a}
% $$

It is helpful to re-express the model as a matrix differential equation, $\mathbf{\dot{u}} = A \mathbf{u}$, where $\mathbf{u} = \left[ u_1, u_2, ... u_N \right]^T$ is the state vector, and $A$ is the state-transition matrix. For a general branched morphology, $A$ will be nearly tridiagonal, with off-diagonal elements corresponding to branch points; matrices in this form are called Hines matrices \citep{Hines_1984}. For the simpler case of an unbranched cable, $A$ is tridiagonal:
$$
A = 
 \begin{bmatrix}
  ~-a_1~  & b_1      & 0         &         & ...              & 0        \\
   a_1    & -b_1-a_2 & b_2       &  0      &                  &          \\
   0      &  a_2     & -b_2-a_3  & ~~~ b_3~~~  & \ddots     & \vdots   \\
   \vdots &  0       & a_3       & \ddots  &                  & 0        \\
          &          & \ddots    &         & -b_{N-2}-a_{N-1} & b_{N-1}  \\
   0      &          & ...       &  0      & a_{N-1}          & ~-b_{N-1}~
 \end{bmatrix}
$$
For both branched and unbranched morphologies, each column of $A$ sums to zero, which reflects conservation of mass within the system. The rank of $A$ is exactly $N-1$ (this can be seen by taking the sum of the first $N-1$ rows, which results in $-1$ times the final row). Thus, the nullspace of $A$ is one-dimensional (red lines in Supp. Fig. 1).

The desired steady-state distribution, $\mathbf{\tilde{u}}$, is an eigenvector that spans the nullspace of $A$. It is simple to show that all other eigenvalues $A$ are negative using the Gershgorin circle theorem; thus, the fixed point described by equations 2 and 4 is stable. The convergence rate is determined by the non-zero eigenvalue with smallest magnitude of $A$. 
%We can increase the convergence rate (while still achieving $\mathbf{\tilde{u}}$) by multiplying all rate constants by some common value. However, this trivial strategy ignores the fact that there are biophysical limits on how fast molecular motors can move.

\subsection*{Stochastic Interpretation of Mass-Action Model}

The mass-action model described in figure 2 is a deterministic dynamical system. However, this determinism only holds in the limit of having a very large number of transported particles. Cargoes with small copy numbers (e.g. mitocondria) will operate in a stochastic regime. We evoke a standard assumption that the system is \textit{ergodic}, meaning that position of one particle averaged over very long time intervals is the same as the ensemble steady-state distribution.

Thus, in addition to modeling how the spatial distribution of cargo changes over time, the mass-action model equivalently models the spatial probability distribution of a single cargo. That is, imagine we track a single cargo and ask its position after a long period of transport. The probability ratio between of finding this particle in any parent-child pair of compartments converges to:
$$
\frac{p_p}{p_c} \bigg|_{ss} = \frac{b}{a}
$$
which mirrors our analysis of the deterministic model.

At steady-state, the number of molecules in any single compartment follows a binomial distribution; the coefficient of variation in each compartment is given by:
$$
\sqrt{\frac{1-p_i^{(ss)}}{n~p_i^{(ss)}}}
$$
This suggests that there are two ways of decreasing noise in the system. First, increasing the number of transported molecules, $n$, decreases the noise by a factor of $1/\sqrt{n}$. Additionally, transport is more reliable to compartments with a high steady-state occupation probability.

\subsection*{Detachment and Degradation Reactions and Quasi-Steady-State Approximation}

Introducing detachment and degradation reactions into the transport model is straightforward. For an arbitrary compartment in a cable, the differential equations become:
$$
\begin{array}{lcl}
\dot{u}_i & = & a_{i-1} u_{i-1} - (a_i+b_{i-1}+c_i) u_i + b_i u_{i+1} \\
\dot{u}_i^\star & = & c_i u_i - d_i u_i^\star
\end{array}
$$

When $a_i,b_i \gg c_i \gg d_i$, then the variables $u_i$ and $u_i^\star$ approach a quasi-steady-state, which we denote $\tilde{u}_i$ and $\tilde{u}_i^\star$. For simplicity we assume $d_i = 0$ in our simulations. We present two strategies for achieving a desired distribution for $\tilde{u}_i^\star$ in figure 4C and 4D. To interpolate between these strategies, let $F$ be a scalar between 0 and 1, and let $\tilde{u}^\star$ be normalized to sum to one. We choose $a_i$ and $b_i$ to achieve:
$$
\tilde{u}_i = F~\tilde{u}_i^\star + (1-F)/N 
$$
along the microtubular network and choose $c_i$ to satisfy
$$
c_i \propto \frac{\tilde{u}_i^\star}{F~\tilde{u}_i^\star + (1-F)/N} 
$$
Here, $N$ is the number of compartments in the model. Setting $F=1$ results in the simulation in figure 4C, while setting $F=0$ results in the simulation in figure 4D. An interpolated strategy is shown in figure 4E ($F=0.3$).

>>>>

\begin{figure}[!tb]
\begin{center}
\includegraphics[width=0.9\columnwidth]{04_sushi_belt.png}
\caption{Multiple strategies for transport in a model including nonuniform microtubular detachment/activation}
\end{center}
\end{figure}

@@

We examined this possibility in a simpler model --- an unbranched cable --- so that we could develop simple heuristics that precisely achieve a desired distribution of cargo while minimizing the convergence time.
%(more complex, \textit{ad hoc}, heuristics could be developed for complex morphologies).
In the reduced model, cargo begins on the left end of the cable and is transported to a number of synaptic sites, each of which is modeled as a double-exponential curve.

\begin{figure}[!tb]
\begin{center}
\includegraphics[width=0.9\columnwidth]{06_heterosynaptic_cable.png}
\caption{Producing speed and specificity for particular stimulation patterns does not extend to other patterns. Cargo begins on the left end and is transported to the right.
\textbf{(A-C)} Steady-state cargo profiles (red) for three stimulation patterns (black arrows) along an unbranched cable.
\textbf{(A)} When the timescale of detachment is sufficiently slow, cargo can be evenly distributed to the synapses regardless of their number and position. Transport parameters were set according to the procedure shown in figure 4D.
\textbf{(B)} When detachment is n\"iavely increased (all rates multiplicatively scaled) a proximal bias in the steady-state distribution of cargo across all stimulation patterns.
\textbf{(C)} Transport rate constants, $a_i$ and $b_i$, were tuned to optimize the distribution of cargo to six equally spaced synapses (top row); detachment rate constants were the same as in panel B. Changing the number of synapses (middle row) or the position of the synapses (bottom row) causes the unequal distribution of cargo to synapses.
}
\end{center}
\end{figure}

@@

We then hand-tuned the transport rate constants to deliver equal cargo to six evenly spaced synaptic sites (top row of Fig. 6C).
Specifically, we increased the anterograde rate constants ($a_i$) and decreased the retrograde rate constants ($b_i$) by a decreasing linear function of position so that $a_N = b_N$ at the right side of the cable (see Methods).
Intuitively, this pushes cargo past the proximal synaptic sites more quickly, canceling out the exponential delivery bias induced by fast detachment.

@@

\subsection*{Microtubule Transport Model (Fig. 1)}
Transport in an unbranched cable (equation 1) leads to the following system of differential equations:
$$
\begin{array}{lcl}
\dot{u}_1 & = & -a_1 u_1 + b_1 u_2 \\
\dot{u}_2 & = & a_1 u_1 - (a_2+b_1) u_2 + b_2 u_3 \\
~ & \vdots & ~ \\
\dot{u}_i & = & a_{i-1} u_{i-1} - (a_i+b_{i-1}) u_i + b_i u_{i+1} \\
~ & \vdots & ~ \\
\dot{u}_N & = & a_{N-1} u_{N-1} - b_{N-1} u_N 
\end{array}
$$
Similar systems result from branched mass-action models. The steady-state relationship in equation 5 can be immediately verified at either end of the cable:
$$
\dot{u}_1 = 0 ~\Rightarrow~ -a_{1} u_{1} + b_{1} u_2 = 0 ~\Rightarrow~ \frac{u_1}{u_2} \Bigg |_{ss} = \frac{b_1}{a_1}
$$
The same procedure can be iteratively applied down the cable to show that equation 5 holds globally. The same result can be shown in a branched tree morphology. Here, each compartment has exactly one parent, but can have multiple children at branch points. Consider a parent-child pair of compartments, $u_p$ and $u_c$, with anterograde/retrograde transport rate constants $a$ and $b$. At steady-state, there is no net flux between $u_c$ and its children:
$$
\dot{u}_c = a u_p - b u_c + \underbrace{\sum_j (b_j u_j - a_j u_c)}_{= 0 \text{ , at ss}}
$$
Thus, we have:
$$
\dot{u}_c = a u_p - b u_c = 0 ~\Rightarrow~ \frac{u_p}{u_c} \Bigg |_{ss} = \frac{b}{a}
$$
To study how the system converges to steady-state, we can re-express the model as a matrix differential equation, $\mathbf{\dot{u}} = A \mathbf{u}$, where $\mathbf{u} = \left[ u_1, u_2, ... u_N \right]^T$ is the state vector, and $A$ is a matrix. For a general branched morphology, the matrix $A$ will be nearly tridiagonal, with off-diagonal elements corresponding to branch points; matrices in this form are called Hines matrices \citep{Hines_1984}. For the simpler case of an unbranched cable, $A$ is tridiagonal:
$$
A = 
 \begin{bmatrix}
  ~-a_1~  & b_1      & 0         &         & ...              & 0        \\
   a_1    & -b_1-a_2 & b_2       &  0      &                  &          \\
   0      &  a_2     & -b_2-a_3  & ~~~ b_3~~~  & \ddots     & \vdots   \\
   \vdots &  0       & a_3       & \ddots  &                  & 0        \\
          &          & \ddots    &         & -b_{N-2}-a_{N-1} & b_{N-1}  \\
   0      &          & ...       &  0      & a_{N-1}          & ~-b_{N-1}~
 \end{bmatrix}
$$
For both branched and unbranched morphologies, each column of $A$ sums to zero, which reflects conservation of mass within the system. The rank of $A$ is exactly $N-1$ (this can be seen by taking the sum of the first $N-1$ rows, which results in $-1$ times the final row). Thus, the nullspace of $A$ is one-dimensional (red lines in Supp. Fig. 1).

The desired distribution, $\mathbf{\tilde{u}}$, is an eigenvector that spans the nullspace of $A$. All other eigenvalues $A$ are negative, which can be shown using the Gershgorin circle theorem. The convergence rate is determined by the non-zero eigenvalue with smallest magnitude of $A$. We can increase the convergence rate (while still maintaining $\mathbf{\tilde{u}}$) by multiplying all rate constants by some common value. However, this trivial strategy ignores the fact that there are biophysical limits on how fast molecular motors can move.

\subsection*{Stochastic Model of Transport}

We model the stochastic transport of a single molecule with a discrete-time Markov chain model. Let the vector $\mathbf{p}^{(k)} = [p_1^{(k)}, p_2^{(k)}, ..., p_N^{(k)}]^T$ denote the probability of finding the molecule in each of the $N$ compartments on the $k$th time step. For example, $p_1^{(0)}$ is the initial probability that the molecule is in the first compartment (which is the soma in our implementation). The state-transition matrix $S$ describes the probabilistic anterograde/retrograde transport of the molecule between neighboring compartments in each discrete time step:
$$
\mathbf{p}^{(k+1)} = S\mathbf{p}^{(k)}
$$
The state-transition matrix is related to the mass-action system matrix $A$ by the following relation:
$$
S = \Delta t A + I
$$
where $\Delta t$ is the length of a time step (assumed to be very small), and $I$ is the identity matrix. The columns of $S$ sum to one, and, since $\Delta t$ is small, all elements of $S$ lie on the interval $[0,1]$. This transformation shifts and scales the eigenvalues of $A$ to lie on the interval $(0,1]$, without modifying the eigenvectors of the system. The steady-state probability distribution is given by the eigenvector of $S$ with eigenvalue 1 (CITE), which is identical to the vector spanning the nullspace of $A$ and therefore the steady-state concentration distribution of the mass-action model.

\subsection{Transport Model With Detachment and Degradation (Fig. 4)}

We can also interpolate between these two strategies to achieve the target distribution. Let $F$ be a scalar between 0 and 1, and let $\tilde{u}^\star$ be normalized to sum to one. We choose $a_i$ and $b_i$ to achieve:
$$
\tilde{u}_i = F~\tilde{u}_i^\star + (1-F)/N 
$$
along the microtubular network and choose $c_i$ to satisfy
$$
c_i \propto \frac{\tilde{u}_i^\star}{F~\tilde{u}_i^\star + (1-F)/N} 
$$
Here, $N$ is the number of compartments in the model. Setting $F=1$ results in the simulation in figure 4C, while setting $F=0$ results in the simulation in figure 4D. Setting $F=0.3$ results in the simulation shown in figure 4E, which uses an interpolated strategy.
